---
title: "Exploring potential urbanization metrics"
format: html
editor_options: 
  chunk_output_type: console
---


(a part of this makes up the Supplementary Material S1)

The goal here is to examine how the different urbanization variables are correlated, to explain the decision to use only "distance to city centroid" in the final text

```{r packages}
library(FactoMineR) # CRAN v2.11 
library(sf)         # CRAN v1.0-20 
library(tidyverse)  # CRAN v2.0.0 

library(here)       # CRAN v1.0.1 
```

We start by loading the coordinates of the sites and of the Rennes urban area centroid (the latter taken from the GHSL UCDB)

```{r load-gpkg}
sites <- read_sf(here("data", "gpkg", "sites.gpkg"))
centroid <- read_sf(here("data", "gpkg", "UCDB_centroid.gpkg")) |>
  st_transform(st_crs(sites))
```

We then load the environmental variables, and filter to keep only sites where _Cornu aspersum_ were found. We then compute the distances to the urban centroid:

```{r load-environment}
# use snail dataset to create filter
with_Cornu <- read_csv(here("data", "snail_phenotype", "snail_shell_size.csv")) |>
  filter(Species == "Cornu_aspersum") |>
  select(SiteID) |>
  distinct()

envt <- read_csv(here("data", "environment", "OCSGE_landcover.csv")) |>
  mutate(has_Cornu = SiteID %in% with_Cornu$SiteID) |>
  mutate(dist_urban = as.numeric(st_distance(sites, centroid)[, 1]))

envt <- filter(envt, has_Cornu)
```

`envt` contains the following:

- `SiteID`: unique site identifier

- `builtup_X`: proportion covered by buildings (CS 1.1.1.1 in OCS-GE, see main text), with X the radius of the circular buffer, in m (from 100 to 1000) 

- `anthropised_X`:proportion covered by all artificialized surfaces including buildings (CS 1.1 in OCS-GE), with X again the radius of the circular buffer

- `has_Cornu`: used above to select sites where _Cornu aspersum_ were found

- `dist_urban`: distance to the centroid of the Rennes urban area as defined in the GHSL UCDB <!--ref-->


# Correlations

## full correlation matrix

```{r corrmatrix-full}
corr_full <- envt |>
  select(starts_with("anthropised"), starts_with("built")) |>
  cor()

mean(corr_full)
range(corr_full)
hist(corr_full)
plot(ecdf(corr_full))
```

## between builtup variables only

```{r corrmatrix-built}
corr_built <- envt |>
  select(starts_with("built")) |>
  cor()

mean(corr_built)
range(corr_built)
hist(corr_built)
plot(ecdf(corr_built))
```

## between builtup variables only

```{r corrmatrix-artificial}
corr_artificial <- envt |>
  select(starts_with("anthropised")) |>
  cor()

mean(corr_artificial)
range(corr_artificial)
hist(corr_artificial)
plot(ecdf(corr_artificial))
```

Very strong correlations, nothing below 0.7 and most > 0.9. Attempting to disentangle them is probably a bit foolish. Let's check this with a PCA

# a PCA of environmental variables

(important: the PCA does **not** include "distance to urban centroid". The point is to look _a posteriori_ how the main axis of the PCA relates to distance)

```{r pca}
ACP <- PCA(envt |> select(-c(SiteID, dist_urban, has_Cornu)))

dimdesc(ACP)
```

PC1 explains most of the variation, all variables are strongly correlated with it and not with any other...

```{r corr_pca_dist}
cor.test(envt$dist_urban, ACP$ind$coord[, 1])
plot(envt$dist_urban, ACP$ind$coord[, 1])
```

... and the distance to urban centroid is also very strongly correlated with PC1, even though it was not included in it.

**Take-home message**: probably not possible to disentangle the spatial scales and urbanisation metrics, best to just use distance to city centroid as single measure of urbanisation
