---
title: "Modelling the effect of urbanisation and shell phenotype on parasite prevalence"
format: html
editor_options: 
  chunk_output_type: console
---


```{r load-packages}
library(brms)      # CRAN v2.22.0 
library(cmdstanr) # https://stan-dev.r-universe.dev v0.9.0 
library(ggtext)    # CRAN v0.1.2 
library(ncf)       # CRAN v1.3-2 
library(patchwork) # CRAN v1.3.0 
library(rptR)      # CRAN v0.9.22 
library(sf)        # CRAN v1.0-20 
library(tidyverse) # CRAN v2.0.0 
library(tidybayes) # CRAN v3.0.7 

library(here)      # CRAN v1.0.1 

options(mc.cores = 4)
```

We first start by importing the raw datasets

```{r import-raw}
sites <- read_sf(here("data", "gpkg", "sites.gpkg"))

centroid <- read_sf(here("data", "gpkg", "UCDB_centroid.gpkg")) |>
  st_transform(st_crs(sites))

snail_size <- read_csv(here("data", "snail_phenotype", "snail_shell_size.csv"))

snail_colour <- read_csv(here("data", "snail_phenotype", "colour_cornu.csv"))

parasites <- read_csv(here("data", "parasites", "combined_parasites_occurrence.csv")) |>
  mutate(
    other_nematodes = as.numeric((nematodes_lung + nematodes_bw_shell_body) > 0),
    sporocerc = as.numeric((sporocysts + cercariae) > 0)
  )
```

(For the description of the columns in `sites`,`centroid`, `snail_size` and `snail_colour`, see the `01-model_size_colour` script)

The `parasites` dataset contains the following columns:

- `SiteID`: a site numbered ID
- `Species`: in the format [genus]_[species]. We only use data from *Cornu aspersum* in this study.
- `SnailID`: unique snail ID, following a [`SiteID`][unique letter]_[first two letters of genus] format
- `nematodes_lung`: (0/1) whether nematodes were found in the snail's lung (multiple potential taxa)         
- `nematodes_bw_shell_body` : (0/1) whether nematodes were found between the snail's shell and its body (multiple potential taxa)      
- `nematodes_genitalia`: (0/1) whether nematodes were found in the snail's genitalia. We identified these as _Nemhelix bakeri_   
- `nematodes_intestine` : (0/1) whether nematodes were found in the snail's intestine (multiple potential taxa)    
- `metacercariae`, `sporocysts` and `cercariae`: (0/1) whether these various trematode larval stages were found in the snail (multiple potential taxa, overwhelmingly _Brachylaima_ sp.)    
- `mites`   : (0/1) whether _Riccardoella_ mites were found
- `nematodes_trapped` and `cercariae_trapped`: whether trematodes or nematodes were found trapped in the shell. There are not used in this study; see Gérard et al. 2023 (doi:10.1051/parasite/2023001): to see them used.

- `sporocerc`: (0/1) Sporocysts and/or cercariae detected. Cercariae are emitted by sporocysts _en masse_ and then leave their host to infect another snail, where they'll become metacercariae. In one individual, we recorded large numbers of cercariae but no sporocysts:

```{r sporocerc}
table(paste("has cercariae?", parasites$cercariae == 1), paste("has sporocysts?", parasites$sporocysts == 1))

parasites |>
  filter(cercariae == 1 & sporocysts == 0)
## individual with SnailID 6L_Co
## looking at the raw parasite files (see 00b-make_occurrence) show it had 50-99 cercariae
```

Given what we know about trematode life cycles, it is extremely unlikely that this many cercariae entered this one snail in search of a new host. It is much, much more likely that this snail harboured sporocysts that were not detected, and that these emitted the cercariae. We therefore used `sporocerc` as synonymous with sporocysts for analyses. 

`other_nematodes`: (0/1) whether `nematodes_bw_shell_body` or `nematodes_lung` is non-zero. We exclude `nematodes_intestine` as only one was found, and we were unable to exclude it was merely endophoretic (Sudhaus 2018, doi:10.25674/4jp6-0v30)       

# Process data

We create the "distance to urban centroid" variable, and create explicit columns for the sites XY coordinates, that will be useful for diagnostics of residual spatial autocorrelation:

```{r distances-urban}
sites$dist_urban <- as.numeric(st_distance(sites, centroid)[, 1])

sites$coordX <- st_coordinates(sites)[, 1]
sites$coordY <- st_coordinates(sites)[, 2]
```


```{r make-table}
tab_parasites <- parasites |>
  left_join(snail_size) |>
  left_join(snail_colour) |>
  left_join(sites) |>
  filter(Species == "Cornu_aspersum") |>
  mutate(size = (diameter_mm_1 + diameter_mm_2) / 2)
```

```{r scaling-predictors}
tab_parasites$size_scaled <- scale(tab_parasites$size)[, 1]
tab_parasites$reflect_scaled <- scale(tab_parasites$avg_reflectance)[, 1]
tab_parasites$sigma_reflect_scaled <- tab_parasites$sigma_reflectance / sd(tab_parasites$avg_reflectance)

sd_size <- sd(tab_parasites$size)
mean_size <- mean(tab_parasites$size)
sd_reflect <- sd(tab_parasites$avg_reflectance)
mean_reflect <- mean(tab_parasites$avg_reflectance)
```

# The model

```{r brm-formulas}
bf_nemhelix <- bf(
  nematodes_genitalia ~
    size_scaled +
    me(reflect_scaled, sigma_reflect_scaled) +
    scale(dist_urban) +
    (1 | p | SiteID),
  family = bernoulli
)

bf_nema_other <- bf(
  other_nematodes ~
    size_scaled +
    me(reflect_scaled, sigma_reflect_scaled) +
    scale(dist_urban) +
    (1 | p | SiteID),
  family = bernoulli
)

bf_meta <- bf(
  metacercariae ~
    size_scaled +
    me(reflect_scaled, sigma_reflect_scaled) +
    scale(dist_urban) +
    (1 | p | SiteID),
  family = bernoulli
)

bf_sporocerc <- bf(
  sporocerc ~
    size_scaled +
    me(reflect_scaled, sigma_reflect_scaled) +
    scale(dist_urban) +
    (1 | p | SiteID),
  family = bernoulli
)
```


```{r model}
mod <- brm(mvbf(bf_nemhelix, bf_nema_other, bf_meta, bf_sporocerc, rescor = FALSE),
  data = tab_parasites,
  backend = "cmdstanr",
  seed = 42,
  iter = 6000, warmup = 3000,
  save_pars = save_pars(group = TRUE, latent = TRUE),
  prior = c(
    set_prior("normal(0,1.5)",
      class = "Intercept",
      resp = c("nematodesgenitalia", "othernematodes", "metacercariae", "sporocerc")
    ),
    set_prior("normal(0,1)",
      class = "b",
      resp = c("nematodesgenitalia", "othernematodes", "metacercariae", "sporocerc")
    ),
    set_prior("normal(0,1)",
      class = "sd",
      resp = c("nematodesgenitalia", "othernematodes", "metacercariae", "sporocerc")
    ),
    ## using an exponential(1) prior for the SDs doesn't change anything
    set_prior("lkj(2)", class = "cor")
  )
)
```

An *example* of posterior predictive checks:

```{r pp-checks}
pp_check(mod, resp = "metacercariae", "bars_grouped", group = "SiteID", ndraws = 1000)
```

We again check the residual spatial autocorrelation (an example, all four parasites similar):

```{r residual-autocor}
residuals_meta <- residuals(mod,
  resp = "metacercariae", re_formula = ~ (1 | SiteID), # type="pearson",
  summary = TRUE
)[, 1]
pp <- spline.correlog(
  x = tab_parasites$coordX / 1000, y = tab_parasites$coordY / 1000,
  z = residuals_meta
)

plot(pp)
```

# Get info for Table and Results

```{r model-summary}
summary(mod)
```

```{r bayes-R2}
bayes_R2(mod, resp = "nematodesgenitalia", re_formula = NA)
bayes_R2(mod, resp = "nematodesgenitalia")

bayes_R2(mod, resp = "othernematodes", re_formula = NA)
bayes_R2(mod, resp = "othernematodes")

bayes_R2(mod, resp = "sporocerc", re_formula = NA)
bayes_R2(mod, resp = "sporocerc")

bayes_R2(mod, resp = "metacercariae", re_formula = NA)
bayes_R2(mod, resp = "metacercariae")
```

# Plots

The model summary shows only two obvious predictor-parasite relationships, so we only plot these two:

```{r plot meta-size}
fits_meta <- tibble(
  size = c(255:410) / 10,
  dist_urban = mean(tab_parasites$dist_urban),
  reflect_scaled = 0,
  sigma_reflect_scaled = mean(tab_parasites$sigma_reflect_scaled)
) |>
  mutate(size_scaled = (size - mean_size) / sd_size) |>
  add_epred_draws(mod, resp = "metacercariae", re_formula = NA) |>
  ungroup()

p_meta <- ggplot(tab_parasites) +
  stat_lineribbon(data = fits_meta, aes(size, .epred), fill = "#9ebcda", .width = 0.95) +
  geom_point(aes(size, metacercariae), col = "#8c96c6", alpha = 0.25) +
  labs(x = "Shell size (mm)", y = "Probability of infection by metacercariae") +
  theme_bw()
p_meta
```

```{r plot-nemhelix-colour}
fits_nemhelix <- tibble(
  avg_reflectance = c(90:230) / 10,
  dist_urban = mean(tab_parasites$dist_urban),
  size_scaled = 0,
  sigma_reflect_scaled = mean(tab_parasites$sigma_reflect_scaled)
) |>
  mutate(reflect_scaled = (avg_reflectance - mean_reflect) / sd_reflect) |>
  add_epred_draws(mod, resp = "nematodesgenitalia", re_formula = NA)

p_nemhelix <- ggplot(tab_parasites) +
  stat_lineribbon(data = fits_nemhelix, aes(avg_reflectance, .epred), fill = "#9ebcda", .width = 0.95) +
  geom_point(aes(avg_reflectance, metacercariae), col = "#8c96c6", alpha = 0.25) +
  labs(x = "Shell reflectance (%)", y = "Probability of infection by _Nemhelix bakeri_") +
  theme_bw() +
  theme(axis.title.y = element_markdown())
p_nemhelix
```


```{r plot}
(p_meta | p_nemhelix)

ggsave(filename = here("R_output", "Fig3.pdf"), width = 8, height = 4)
```
